
---
output: html_document
editor_options:
 chunk_output_type: console
---

# Read "recipes", read IRS files, and save preliminary version of target data

This chapter reads the manually created "target recipes" xlsx file and reads the IRS xls* files identifed within.

It then saves them as potential_targets_preliminary.rds, to be used for further investigation and testing to be sure they seem correct.

## Setup

```{r}
#| label: setup

GITROOT <- rprojroot::find_root(rprojroot::has_file(".git/index"))
QROOT <- fs::path(GITROOT, "tmd", "national_targets", "qmd")
source(fs::path(QROOT, "setup.R"))
source(fs::path(QROOT, "R", "functions_excel.R")) # drop from setup if not used elsewhere

targfn <- "target_recipes_v3.xlsx"

# note - will load the functions that help with reading excel files
# also loads QROOT, DATADIR, and targfn

```

## Functions for reading the excel file

```{r}
#| label: get-rowmap

get_rowmap <- function(tab, DATADIR, targfn) {
  # reads the target recipes xlsx file to
  # get start and end row for key data for each year of a particular IRS spreadsheet
  # from its associated mapping tab in the recipes file
  # targfn is the targets filename
  sheet <- paste0(tab, "_map")
  readxl::read_excel(
    fs::path(DATADIR, targfn),
    sheet = sheet,
    range = "E1:H3"
  ) |>
    tidyr::pivot_longer(
      -indextype,
      names_to = "year",
      values_to = "xlrownum"
    ) |>
    dplyr::mutate(
      table = tab,
      year = as.integer(year),
      xlrownum = as.integer(xlrownum)
    ) |>
    dplyr::select(table, year, indextype, xlrownum) |>
    dplyr::arrange(table, year, desc(indextype))
}
```


```{r}
#| label: get-colmap

get_colmap <- function(tab, DATADIR, targfn) {
  # reads the target_recipes.xlsx file to
  # get columns of interest for each year of a particular IRS spreadsheet,
  # from its associated mapping tab in the recipes file

  # assumes DATADIR and targfn (targets filename) are in the environment
  sheet <- paste0(tab, "_map")
  col_map <- readxl::read_excel(
    path(DATADIR, targfn),
    sheet = sheet,
    skip = 3,
    col_types = "text"
  ) |>
    filter(!is.na(vname)) |>
    select(
      vname,
      vtype,
      subgroup,
      marstat,
      vdescription,
      `2015`,
      `2021`,
      `2022`
    ) |>
    pivot_longer(
      -c(vname, vtype, subgroup, marstat, vdescription),
      names_to = "year",
      values_to = "xlcolumn"
    ) |>
    filter(!is.na(xlcolumn)) |> # only keep variable-year combos that we have identified in the excel file
    mutate(
      table = tab,
      year = as.integer(year),
      xl_colnumber = xlcol_to_num(xlcolumn)
    ) |>
    select(
      table,
      year,
      vname,
      vtype,
      subgroup,
      marstat,
      xl_colnumber,
      xlcolumn,
      vdescription
    ) |>
    arrange(table, vtype, year, vname, subgroup, marstat, xl_colnumber)
  col_map
}

```


```{r}
#| label: get-irs-table

# test data for get_irsdata
# fname <- "15in11si.xls"; year <- 2015
# startrow <- 10; endrow <- 29; max_xl_colnumber <- 7; column_letters <- list(c("A", "B", "D", "G"))
#  vnames <- list(c("incrange", "x2", "x3", "x4"))

get_irs_table <- function(
  fname,
  year,
  startrow,
  endrow,
  max_xl_colnumber,
  column_letters
) {
  # read a single file for a single year
  # print(fname)
  fpath <- fs::path(DATADIR, year, fname)

  # read relevant rows, but all columns from the first column through the last in columns
  #   we cannot read just the desired columns, due to limitations of read_excel

  df1 <- read_excel(
    fpath,
    sheet = 1,
    range = cellranger::cell_limits(
      c(startrow, 1),
      c(endrow, max_xl_colnumber)
    ),
    col_names = xlnum_to_col(1:max_xl_colnumber),
    col_types = "text"
  )

  # keep desired columns, substitute the passed-in column names for letters, and pivot
  df2 <- df1 |>
    select(all_of(unlist(column_letters))) |>
    rename(incrange = 1) |>
    # setNames(unlist(vnames)) |>
    mutate(xlrownum = startrow:endrow, incsort = row_number()) |>
    pivot_longer(
      cols = -c(incsort, xlrownum, incrange),
      names_to = "xlcolumn",
      values_to = "ptarget"
    ) |>
    mutate(ptarget = as.numeric(ptarget)) |> # every column had better be a number stored as text!
    select(incsort, xlrownum, incrange, xlcolumn, ptarget) |>
    arrange(xlcolumn, incsort)
  df2
}

```

## Read the target_recipes tab that defines IRS tables of interest

Get and display information on IRS tables we plan to use.


```{r}
#| label: define-display-tables-needed
#| output: true
#| column: page

needed_irs_tables <- read_excel(
  fs::path(DATADIR, targfn),
  sheet = "irs_downloads"
)

fnote <- md(
  "IRS filenames are constructed by prepending a 2-digit year to the base filename. For example, the base file name for Table 1.1 (sources of income -- si) is *in11si.xls*; therefore the Table 1.1 file name for 2022 is *22in11si.xls*."
)

needed_irs_tables |>
  gt::gt() |>
  gt::tab_header("Key IRS tables that contain potential targets") |>
  gt::cols_label(
    fname_base = "Base file name"
  ) |>
  gt::tab_footnote(fnote, locations = cells_column_labels(fname_base)) |>
  gt::tab_options(table.width = gt::pct(100))

```


Create table metadata file with IRS file names for each year of interest (base file name prepended by 2-digit year).

```{r}
#| label: get-recipes
#| output: false

# table_meta has information about each Excel file we want to download
table_meta <- expand_grid(
  year = c(2015, 2021:2022) |>
    as.integer(),
  needed_irs_tables
) |>
  mutate(
    fname = paste0(str_sub(year, 3, 4), fname_base),
    upath = paste0(baseurl, fname)
  ) |>
  relocate(fname, .after = table)

glimpse(table_meta)
table_meta |>
  select(1:4)

```


## Read IRS spreadsheets and save data frame of targets as csv file


```{r}
#| label: read-irs-spreadsheets
#| output: false

# create a csv file (tibble) with the following columns:
#   table -- tab11, tab12, tab14, tab21
#   year (2015, 2021, 2022)
#   fname
#   table_description
#   vtype (stub, count, amount)
#   subgroup (filers, TBD)
#   marstat (all, mfjss, mfs, single, hoh)
#   vname (name we are using for an IRS variable)
#   vdescription
#   incsort ()
#   incrange (text description, may vary from file to file)
#   xlrownum
#   xlcolumn
#   ptarget (potential target)

# combine table metadata, row specs, and column specs to get the specs needed to read the data files and get the data
# after getting data merge back against variable specs (used to create column specs) to get variable names and other variable info
tabs <- c("tab11", "tab12", "tab14", "tab21")

table_meta

# get start and end rows for each file of interest
table_rowspecs <- tabs |>
  purrr::map(\(tab) get_rowmap(tab, DATADIR, targfn)) |>
  list_rbind() |>
  pivot_wider(names_from = indextype, values_from = xlrownum)
table_rowspecs

table_variable_specs <- tabs |>
  purrr::map(\(tab) get_colmap(tab, DATADIR, targfn)) |>
  list_rbind() |>
  arrange(table, year, xl_colnumber)
# TODO: make sure we don't have any repeated column numbers within a table -- that would indicate an error in the recipe file

table_colspecs <- table_variable_specs |>
  dplyr::summarise(
    nvars = n(),
    max_xl_colnumber = max(xl_colnumber),
    column_letters = list(xlcolumn),
    .by = c(table, year)
  )
skimr::skim_without_charts(table_colspecs)
if (anyNA(table_colspecs)) {
  stop("Found NA values in the tibble table_colspecs.")
}

table_specs <- table_meta |>
  filter(table %in% tabs) |>
  select(table, year, fname, table_description, table_type) |>
  left_join(table_rowspecs, by = join_by(year, table)) |>
  left_join(table_colspecs, by = join_by(year, table))
# now we have what we need to get the data from each table

table_data <- table_specs |>
  mutate(
    targets = get_irs_table(
      fname,
      year,
      startrow,
      endrow,
      max_xl_colnumber,
      column_letters
    ) |>
      list(),
    .by = fname
  )

table_data
table_data$targets[1]

ptargets_for_checking <- table_data |>
  unnest(cols = targets) |>
  # bring in variable specs for each variable
  left_join(table_variable_specs, by = join_by(table, year, xlcolumn)) |>
  select(
    year,
    vname,
    table,
    vdescription,
    vtype,
    subgroup,
    marstat,
    incsort,
    incrange,
    ptarget,
    everything()
  ) |>
  arrange(year, vname, table, vdescription, vtype, subgroup, marstat, incsort)

# TODO: do some checks on the data

# check for identical
# sha256sum potential_targets_original.csv potential_targets.csv

```


## Save preliminary target data

```{r}

# get a slimmed down table for saving
ptargets <- ptargets_for_checking |>
  arrange(year, table, vname, vtype, subgroup, marstat, incsort) |>
  mutate(
    rownum = row_number(),
    idbase = paste(
      table,
      year,
      vname,
      vtype,
      subgroup,
      marstat,
      sep = "-"
    ),
    # xlrownum = str_pad(xlrownum, width = 2, side = "left", pad = "0"),
    xlcell = paste0(xlcolumn, xlrownum)
  ) |>
  select(
    rownum,
    idbase,
    year,
    table,
    vname,
    vtype,
    vdescription,
    subgroup,
    marstat,
    incsort,
    incrange,
    ptarget,
    fname,
    xlcell #,
    # xlcolumn,
    # xlrownum
  )

```

```{r}
#| label: save-raw-irs-data
#| output: false

saveRDS(ptargets, fs::path(DATADIR, "potential_targets_preliminary.rds"))

ptargets |>
  write_csv(fs::path(DATADIR, "potential_targets_preliminary.csv"))

```
